Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-06-13 doc/invt.tex
invt.tex: remove appendix on Gamma

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $u_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the resulting distribution, $\pi_i(t)$,
the real deviation of the bias potential
depends only on the difference between $v_i(t)$
and a baseline value, $\bav{v}(t)$,
for average growth\cite{
dama2014}.
%
We define a bin-average-shifted bias potential as
%
\begin{equation}
  v_{*i}(t) \equiv v_i(t) - \bav{v}(t)
  ,
\label{eq:x_def}
\end{equation}
%
where the bin average, $\bav{v}(t)$, is defined as
%
\begin{equation}
  \bav{v}(t)
  \equiv \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vav_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$
(with $\rho_i \ge 0$ and $\sum_{i = 1}^n \rho_i = 1$).
%
Below we will use the notations, $f_{*i}$ and $\bav{f}$,
for an arbitrary variable, $f_i$.
%
Note that the distribution, $\pmb\rho$,
can be independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$,
and chosen for convenience.\footnote{In
well-tempered metadynamics,
$\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
  barducci2008, dama2014}:
$\ln p_i = \ln {p^*_i}/(1+\Delta) + \mathrm{const.}$ for some positive $\Delta$,
whereas $\pmb\rho$ is flat.}

@@

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation\footnote{This
approximation requires us to shift the origin of time by $1$ in some cases.}
%
\begin{equation}
  \dot u_i(t)
  \equiv
  \frac{ d u_i(t) } { dt }
  \approx
  \alpha(t) \, \frac{ h_i(t) } { p_i }
  \equiv
  \alpha(t) \, \xi_i(t)
  ,
  \label{eq:ut_diffeq}
\end{equation}
%
with
%
$h_i(t) = \delta_{i, i(t)}$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1$
for the bin $i(t)$ at time $t$
or zero otherwise,
and we have defined a relative histogram
$\xi_i(t) \equiv h_i(t) /p_i$
in the last step.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.
%
Deducting the bin average [cf. Eq. \eqref{eq:x_def}],
we get
%
\begin{equation}
  \dot v_{*i}(t)
  \approx
  \alpha(t) \, \xi_{*i}(t)
  .
  \label{eq:vt_diffeq}
\end{equation}

We can split the histogram into
a deterministic averaged part, $\langle \xi_i(t) \rangle$,
and a random fluctuation part, $\zeta_i(t)$,
%
\begin{equation}
  \xi_i(t) =
  \langle \xi_i(t) \rangle
  +
  \zeta_i(t)
  .
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an average in an ensemble consists of
many copies of similar simulations
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.
%
For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period,
and we may assume a quasi-equilibrium sampling process
such that
%with the deterministic part specified by Eq. \eqref{eq:pi_p_V}:
%
\begin{align}
  \langle \xi_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { p_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { p_i }
  =
  \frac{                       e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n p_j \, e^{-v_{*j}(t)} }
  \notag
  \\
  &\approx
  1 - v_{*i}(t) + \sum_{j = 1}^n p_j \, v_{*j}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n p_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n p_j \, v_{*j} }
.
$$
}%
%
Deducting the bin average, we get
%
\begin{align}
  \langle \xi_{*i}(t) \rangle
  \equiv
  \langle \xi_i(t) \rangle -
  \sum_{j=1}^n \rho_j \langle \xi_j(t) \rangle
  &\approx
  - v_{*i}(t)
  .
  \label{eq:sh_ave}
\end{align}

We will approximate the histogram fluctuation part, $\zeta_i(t)$,
hence its bin-average-shifted version, $\zeta_{*i}(t)$,
as white noise such that
\begin{equation}
  \langle \zeta_{*i}(t) \, \zeta_{*i}(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:G_def}
\end{equation}
for some $G_i$,
and $\delta(t)$ is the Dirac's delta function.
%
We expect the approximation to hold for simulation
much longer than the autocorrelation time.
%
For non-white noise,
the equivalent value of $G_i$ can be approximately obtained as
$\sum_{t = -\infty}^\infty \langle \zeta_{*i}(t) \, \zeta_{*i}(0) \rangle$.

@@

The error depends implicitly on the curve, $\alpha(t)$,
or, equivalently, the curve, $q(t)$,
via $v_{*i}(T)$, and
our aim is to find the $q(t)$
that minimizes this error.
%
We now fix the endpoint value, $q(T)$,
and vary the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to varying the curve $y(q(t))$
with fixed endpoints,
$y\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$y\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:ET_average} is fixed during
the process,
while the second term is the action of a free particle.
Thus, the velocity of $y(q(t))$ should be a constant:
%
\begin{equation}
  \dot y\bigl( q(t) \bigr) = c
  .
\label{eq:dydt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:y_def}, we have
%
\begin{equation}
  e^{ q(t) - q(T) }
  =
  c \, (t + t_0)
  =
  \frac{ t + t_0 }
       { T + t_0 }
  ,
\label{eq:expqt}
\end{equation}
where $t_0$ is some constant,
and the other constant, $c$, has been determined
from the $t = T$ case
in the second step.
%
Taking the logarithm,
and differentiating both sides
with respect to $t$,
we get the inverse-time schedule
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }
  ,
\label{eq:alpha_invt1}
\end{equation}
%
and Eq. \eqref{eq:alpha_invt}
is the special case of $t_0 = 0$.
%

We now determine the optimal value of $q(T)$,
which is fixed in the above process.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
from the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
By using Eq. \eqref{eq:alpha_invt} in Eq. \eqref{eq:ET_average},
we find the final total error to be
%
$E(T) = [ E(0) \, t_0^2 + \Gamma \, T ]/(t_0 + T)^2$,
%
which is minimized if
%
\begin{equation}
t_0 = \frac{ \Gamma }
           { E(0) }
           ,
  \label{eq:t0_singlebin}
\end{equation}
%
and at this point, the error is given by
%
\begin{equation}
  E(T)
  =
  \frac{ \Gamma } { T + t_0 }
  .
  \label{eq:Emin_singlebin}
\end{equation}
%
This error is roughly equal to
the overall histogram flatness\cite{zhou2005, zhou2008}
and thus it can be conveniently estimated
(see also Appendix \ref{sec:flatness}).

@@

%To simplify the ensuing discussion,
Second, we limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  ,
  \label{eq:w_detailedbalance}
\end{equation}
%
for the error weighting distribution, $\pmb \rho$,
introduced in Eq. \eqref{eq:vav_def}.
%
This requires the updating matrix, $\mathbf w$
to be symmetric
for a uniform $\pmb\rho$,
or more generally,
the scaled matrix, $\sqrt{ \rho_i/\rho_j } \, w_{ij}$,
to be symmetric.
%
This symmetry allows the diagonalization
of $\mathbf w$ with a set of
eigenvectors, $\phi_{ki}$,
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 0}^{n - 1}
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, \rho_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { \rho_i }
  &=
  \delta_{kl}
  ,
  \label{eq:eig_orthonormal_rows}
\end{align}
%
where we have set the index of the first eigenmode to $0$
instead of $1$ for later convenience.

The above symmetry implies the existence of
a left eigenvector, $\pmb \rho$,
corresponding to the uniform right eigenvector
associated with Eq. \eqref{eq:w_sumj},
with eigenvalue $1$:
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j
  .
  \notag
  %\label{eq:w_balance}
\end{equation}
%
We will label this eigenvector as the first one,
and
%
\begin{equation}
  \phi_{0i} = \rho_i,
\label{eq:eigenmode0}
\end{equation}
%
with $\lambda_0 = 1$.
%
Clearly, Eq. \eqref{eq:eigenmode0}
satisfies the normalization condition
given by Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 0$,
while the orthogonality demands
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \delta_{k0}
  .
\label{eq:ortho1}
\end{equation}
%
We may write the updating matrix in $\phi_{ki}$'s
as\cite{bussi2006}
%
\begin{equation}
  w_{ij}
  =
  \frac{1}{\rho_i} \sum_{k=0}^{n - 1}
  \lambda_k \, \phi_{ki} \, \phi_{kj}
  .
  \notag
\end{equation}

@@

We can further define
a generalized Fourier transform, $\mathcal{F}$,
for variable, $A_i$,
%
\begin{equation}
  {\tilde A}_k
  \equiv \mathcal{F}[A_i]_k
  \equiv \sum_{i = 1}^n \phi_{ki} \, A_i
  ,
  \notag
  %\label{eq:gft_def}
\end{equation}
%
and the inverse transform is
%
\begin{equation}
  A_i =
  \frac{1}{\rho_i} \sum_{k = 1}^n \phi_{ki} \, \tilde{A}_k
  .
  \notag
  %\label{eq:gft_inv}
\end{equation}
%
Using this definition,
we can rewrite Eq. \eqref{eq:ortho1} as
$\mathcal F[1]_k = \delta_{k0}$,
and for the bin average of any quantity, $f$, we have
$\bav{f}= \mathcal F[f_i]_0 = \tilde f_0$,
so that
\begin{equation}
  \tilde f_{*k}
  = \tilde f_k - \bav{f} \, \delta_{k0}
  =
  \begin{dcases}
    0           & k = 0, \\
    \tilde f_k  & k > 0,
  \end{dcases}
  \label{eq:ftstar}
\end{equation}
%
which means that the only effect
of the bin-average deduction is to
annihilate the first mode, $\tilde f_0$.

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
$$
{\tilde v}_k(t + 1) =
{\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
{\tilde \xi}_k(t)
,
$$
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde \xi}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \zeta}_k(t) \bigr]
  ,
  \label{eq:ut_diffeq_mbin}
\end{equation}
%
where we have removed the bin average,
and used the transformed version of Eq. \eqref{eq:sh_ave}
[the subscript ``$_*$'' is dropped by Eq. \eqref{eq:ftstar}].
In the second step, we have used the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.

@@

In analogy to Eq. \eqref{eq:ET_average},
%using Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:xt2_sum},
we can write the error after a period, $T$,
as a sum of two contributions,
%
\begin{equation}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T),
  \label{eq:error_tot}
\end{equation}
%
where the residual error
(from the decay of the initial error)
is given by
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 1}^{n-1}
    \left\langle
      {\tilde v}_k^2(0)
    \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
is given by
%
\begin{align}
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 1}^{n-1}
  \Gamma_k \, \dot y_k^2\bigl( q(t) \bigr) \, dt
  ,
\label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  y_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\kappa_k(t) \equiv \bigl\langle {\tilde \zeta}_k(0)
\, {\tilde \zeta}_k(t) \bigr\rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty \kappa_k(t)
  = \sum_{t = -\infty}^\infty
  \left\langle {\tilde \zeta}_k(0)
            \, {\tilde \zeta}_k(t) \right\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
We may understand Eq. \eqref{eq:Gamma_sum}
as an attempt of modeling
a general histogram fluctuation, ${\tilde \zeta}_k(t)$,
as an equivalent white noise
so that the theory can be applicable to
a general sampling process.
%
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
For example, we expect large values of $\Gamma_k$
from eigenmodes corresponding to
persistent long-range motion
in sampling complex or glassy systems.
%
In Appendix \ref{sec:Gamma},
we discuss several ways of estimating $\Gamma_k$'s.
%

@@

To model the initial error for Eq. \eqref{eq:error_res},
we will assume that
the preliminary adaptive FDS sampling has been
performed under a constant updating magnitude for $T_0$, steps.
%
If the starting bias potential is zero, $u_i = 0$,
one can show that
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  + \epsilon_k
  ,
  \label{eq:xt2_eql1}
\end{equation}
%
where
$\epsilon_k \equiv \tilde u_k^2 \, e^{-2\, \lambda_k \, a_0 \, T_0}$
and if no eigenvalue is zero,
then we can ignore the $\epsilon_k$ term
in the case of $T_0 \to \infty$,
%
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  .
  \label{eq:xt2_eql}
\end{equation}
%
\note{
Consider a long period, $T_0$, under a fixed schedule,
$\alpha(t) = a_0$,
the residual error becomes negligible, and
the $k$th component of the asymptotic error
is given by
%
\begin{align*}
  \left\langle
    {\tilde v}_k^2(T_0)
  \right\rangle
  =
  \int_0^{T_0}
    \Gamma_k \, (\lambda_k \, a_0)^2 \,
      e^{ 2 \, \lambda_k \, a_0 \, (t - T_0) }
    \, dt
  \stackrel{ T_0 \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  .
\notag
%\label{eq:error_eql}
\end{align*}
}
%

>>>>

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $u_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the resulting distribution, $\pi_i(t)$,
the real deviation of the bias potential
depends only on the difference between $v_i(t)$
and a baseline value, $\bav{v}(t)$,
for average growth\cite{
dama2014}.
%
We define a bin-average-shifted bias potential as
%
\begin{equation}
  v_{*i}(t) \equiv v_i(t) - \bav{v}(t)
  ,
\label{eq:x_def}
\end{equation}
%
where the bin average, $\bav{v}(t)$, is defined as
%
\begin{equation}
  \bav{v}(t)
  \equiv \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vav_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$
(with $\rho_i \ge 0$ and $\sum_{i = 1}^n \rho_i = 1$).
%
Below we will use the notations, $f_{*i}$ and $\bav{f}$,
for an arbitrary variable, $f_i$.
%
Note that the distribution, $\pmb\rho$,
can be independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$,
and chosen for convenience.\footnote{Alternatively,
we may define $v_{*i}(t)$ such that
$\rho_i \, e^{-v_{*i}(t)}$ is a normalized distribution,
i.e.,
$$
v_{*i}(t) \to -\log\left(
  \frac{ (p^*_i/p_i) \, e^{ -u_i(t) } } 
  { \sum_{j=1}^n \rho_j \, (p^*_j/p_j) \, e^{ -u_j(t) } } 
\right).
$$
One can readily show that the two definitions are equivalent
for small $v_{*i}$.
%
In well-tempered metadynamics,
$\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
  barducci2008, dama2014}:
$\ln p_i = \ln {p^*_i}/(1+\Delta) + \mathrm{const.}$ for some positive $\Delta$,
whereas $\pmb\rho$ is flat.}

@@

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation\footnote{In
going to the continuous time setup,
we find it more convenient to shift the origin of time by $-1$,
e.g. the sum $\sum_{i=1}^T$ is mapped to the integral $\int_0^T dt$.}
%
\begin{equation}
  \dot u_i(t)
  \equiv
  \frac{ d u_i(t) } { dt }
  \approx
  \alpha(t) \, \frac{ h_i(t) } { p_i }
  \equiv
  \alpha(t) \, \xi_i(t)
  ,
  \label{eq:ut_diffeq}
\end{equation}
%
with
%
$h_i(t) = \delta_{i, i(t)}$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1$
for the bin $i(t)$ at time $t$
or zero otherwise,
and we have defined
$\xi_i(t) \equiv h_i(t) /p_i$.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.
%
Deducting the bin average [cf. Eq. \eqref{eq:x_def}],
we get
%
\begin{equation}
  \dot v_{*i}(t)
  \approx
  \alpha(t) \, \xi_{*i}(t)
  .
  \label{eq:vt_diffeq}
\end{equation}

We can split the histogram into
a deterministic averaged part, $\langle \xi_i(t) \rangle$,
and a random fluctuation part, $\zeta_i(t)$,
%
\begin{equation}
  \xi_i(t) =
  \langle \xi_i(t) \rangle
  +
  \zeta_i(t)
  .
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an average in an ensemble consisting of
many copies of similar simulations
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.
%
For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period,
and we may assume a quasi-equilibrium sampling process
such that
%with the deterministic part specified by Eq. \eqref{eq:pi_p_V}:
%
\begin{align}
  \langle \xi_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { p_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { p_i }
  =
  \frac{                       e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n p_j \, e^{-v_{*j}(t)} }
  \notag
  \\
  &\approx
  1 - v_{*i}(t) + \sum_{j = 1}^n p_j \, v_{*j}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n p_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n p_j \, v_{*j} }
.
$$
}%
%
Deducting the bin average, we get
%
\begin{align}
  \langle \xi_{*i}(t) \rangle
  \equiv
  \langle \xi_i(t) \rangle -
  \sum_{j=1}^n \rho_j \langle \xi_j(t) \rangle
  &\approx
  - v_{*i}(t)
  .
  \label{eq:sh_ave}
\end{align}

We will approximate the histogram fluctuation part, $\zeta_i(t)$,
hence its bin-average-shifted version, $\zeta_{*i}(t)$,
as white noise such that
\begin{equation}
  \langle \zeta_{*i}(t) \, \zeta_{*i}(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:G_def}
\end{equation}
for some $G_i$,
and $\delta(t)$ is the Dirac's delta function.
%
We expect the approximation to hold for simulation
much longer than the autocorrelation time.
%
%For non-white noise,
%the equivalent value of $G_i$ can be approximately obtained as
%$\sum_{t = -\infty}^\infty \langle \zeta_{*i}(t) \, \zeta_{*i}(0) \rangle$.

@@

The error depends implicitly on the curve, $\alpha(t)$,
or, equivalently, the curve, $q(t)$,
via $v_{*i}(T)$, and
our aim is to find the $q(t)$
that minimizes this error.
%
We now fix the endpoint value, $q(T)$,
and vary the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to varying the curve $y(q(t))$
with fixed endpoints,
$y\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$y\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:ET_average} is fixed during
the process,
while the second term is the action of a free particle.
Thus, the velocity of $y(q(t))$ should be a constant:
%
\begin{equation}
  \dot y\bigl( q(t) \bigr) = c
  .
\label{eq:dydt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:y_def}, we have
%
\begin{equation}
  e^{ q(t) - q(T) }
  =
  c \, (t + t_0)
  =
  \frac{ t + t_0 }
       { T + t_0 }
  ,
\label{eq:expqt}
\end{equation}
where $t_0$ is some constant,
and $c$ has been determined
from the $t = T$ case
in the second step.
%
Taking the logarithm,
and differentiating both sides
with respect to $t$,
we get the inverse-time schedule
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }
  ,
\label{eq:alpha_invt1}
\end{equation}
%
and Eq. \eqref{eq:alpha_invt}
is the special case of $t_0 = 0$.
%

We now determine the optimal value of $q(T)$,
which is fixed in the above process.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
from the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
By using Eq. \eqref{eq:alpha_invt} in Eq. \eqref{eq:ET_average},
we find the final total error to be
%
$E(T) = [ E(0) \, t_0^2 + \Gamma \, T ]/(t_0 + T)^2$,
%
which is minimal at
%
\begin{equation}
t_0 = \frac{ \Gamma }
           { E(0) }
           ,
  \label{eq:t0_singlebin}
\end{equation}
%
and the minimum is
%
\begin{equation}
  E(T)
  =
  \frac{ \Gamma } { T + t_0 }
  .
  \label{eq:Emin_singlebin}
\end{equation}
%
This error is roughly equal to
the average histogram fluctuation\cite{zhou2005, zhou2008}
(see also Appendix \ref{sec:hfluc}).

@@

%To simplify the ensuing discussion,
Second, we limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  ,
  \label{eq:w_detailedbalance}
\end{equation}
%
for the error weighting distribution, $\pmb \rho$,
introduced in Eq. \eqref{eq:vav_def}.
%
This requires the scaled updating matrix,
$\sqrt{ \rho_i/\rho_j } \, w_{ij}$,
to be symmetric,
and the symmetry allows the diagonalization
of $\mathbf w$ with a set of
eigenvectors, $\phi_{ki}$,
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 0}^{n - 1}
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, \rho_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { \rho_i }
  &=
  \delta_{kl}
  ,
  \label{eq:eig_orthonormal_rows}
\end{align}
%
where we have set the index of the first eigenmode to $0$
instead of $1$ for later convenience.

The above symmetry implies the existence of
a left eigenvector, $\pmb \rho$,
corresponding to the uniform right eigenvector
associated with Eq. \eqref{eq:w_sumj},
with eigenvalue $1$:
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j
  .
  \notag
  %\label{eq:w_balance}
\end{equation}
%
We will label this eigenvector as the first one,
and
%
\begin{equation}
  \phi_{0i} = \rho_i,
\label{eq:eigenmode0}
\end{equation}
%
with $\lambda_0 = 1$.
%
Clearly, Eq. \eqref{eq:eigenmode0}
satisfies the normalization condition
given by Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 0$,
while the orthogonality demands
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \delta_{k0}
  .
\label{eq:ortho1}
\end{equation}
%
%We may write the updating matrix in $\phi_{ki}$'s
%as\cite{bussi2006}
%%
%\begin{equation}
%  w_{ij}
%  =
%  \frac{1}{\rho_i} \sum_{k=0}^{n - 1}
%  \lambda_k \, \phi_{ki} \, \phi_{kj}
%  .
%  \notag
%\end{equation}

@@

We can further define
a generalized Fourier transform, $\mathcal{F}$,
for variable, $A_i$,
%
\begin{equation}
  {\tilde A}_k
  \equiv \mathcal{F}[A_i]_k
  \equiv \sum_{i = 1}^n \phi_{ki} \, A_i
  ,
  \notag
  %\label{eq:gft_def}
\end{equation}
%
and the inverse transform is
%
\begin{equation}
  A_i =
  \frac{1}{\rho_i} \sum_{k = 1}^n \phi_{ki} \, \tilde{A}_k
  .
  \notag
  %\label{eq:gft_inv}
\end{equation}
%
Using this definition,
we can rewrite Eq. \eqref{eq:ortho1} as
$\mathcal F[1]_k = \delta_{k0}$,
%
and Eq. \eqref{eq:eigenmode0} means that
the bin average of any quantity, $f$,
is the first mode of the Fourier transform:
$\bav{f}= \mathcal F[f_i]_0 = \tilde f_0$.
%
So
\begin{equation}
  \tilde f_{*k}
  = \tilde f_k - \bav{f} \, \delta_{k0}
  =
  \begin{dcases}
    0           & k = 0, \\
    \tilde f_k  & k > 0,
  \end{dcases}
  \label{eq:ftstar}
\end{equation}
%
which means that the only effect
of the bin-average deduction on $\tilde f_k$ is to
annihilate the first mode, $\tilde f_0$.

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
$$
{\tilde v}_k(t + 1) =
{\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
{\tilde \xi}_k(t)
,
$$
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde \xi}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \zeta}_k(t) \bigr]
  ,
  \label{eq:ut_diffeq_mbin}
\end{equation}
%
where we have removed the bin average,
and used the transformed version of Eq. \eqref{eq:sh_ave}
[but the subscript ``$_*$'' is dropped by Eq. \eqref{eq:ftstar}].
In the second step, we have used the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.

@@

In analogy to Eq. \eqref{eq:ET_average},
%using Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:xt2_sum},
we can write the error after a period, $T$,
as a sum of two contributions,
%
\begin{equation}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T),
  \label{eq:error_tot}
\end{equation}
%
where the residual error
(from the decay of the initial error)
is given by
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 1}^{n-1}
    \left\langle
      {\tilde v}_k^2(0)
    \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
is given by
%
\begin{align}
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 1}^{n-1}
  \Gamma_k \, \dot y_k^2\bigl( q(t) \bigr) \, dt
  ,
\label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  y_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\kappa_k(t) \equiv \bigl\langle {\tilde \zeta}_k(0)
\, {\tilde \zeta}_k(t) \bigr\rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty \kappa_k(t)
  = \sum_{t = -\infty}^\infty
  \left\langle {\tilde \zeta}_k(0)
            \, {\tilde \zeta}_k(t) \right\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
Equation \eqref{eq:Gamma_sum} provides a way
of approximately modeling
a general histogram fluctuation, ${\tilde \zeta}_k(t)$,
as an equivalent white noise.
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
In Appendix \ref{sec:Gamma_measure},
we discuss a way of estimating $\Gamma_k$
from an FDS simulation with a constant magnitude.
%

@@

To model the initial error for Eq. \eqref{eq:error_res},
we will assume that
the preliminary adaptive FDS sampling has been
performed under a constant updating magnitude for $T_0$ steps.
%
If the starting bias potential is zero, $u_i = 0$,
one can show that
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  + \epsilon_k
  ,
  \label{eq:xt2_eql1}
\end{equation}
%
where
$\epsilon_k \equiv \tilde u_k^2 \, e^{-2\, \lambda_k \, a_0 \, T_0}$
and if no eigenvalue is zero,
then we can ignore the $\epsilon_k$ term
in the case of $T_0 \to \infty$,
%
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  .
  \label{eq:xt2_eql}
\end{equation}
%
\note{
Consider a long period, $T_0$, under a fixed schedule,
$\alpha(t) = a_0$,
the residual error becomes negligible, and
the $k$th component of the asymptotic error
is given by
%
\begin{align*}
  \left\langle
    {\tilde v}_k^2(T_0)
  \right\rangle
  =
  \int_0^{T_0}
    \Gamma_k \, (\lambda_k \, a_0)^2 \,
      e^{ 2 \, \lambda_k \, a_0 \, (t - T_0) }
    \, dt
  \stackrel{ T_0 \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  .
\notag
%\label{eq:error_eql}
\end{align*}
}
%
